Introduction

### 2020-08-28 text/discussion.Rmd
cut all text to conform to Cell's 45k character limit

One of the biggest hurdles in cancer research is the sparsity of data; ~20,000
protein-coding genes is comparable with the number of tumor samples, even with
multiple mutations per sample. This also does not account for the fact that
~30,000 non-coding genes and hundreds of thousands of non-coding regulatory
elements exist with oncogenic or tumor suppressor activity [@deesMuSiC2012;
@tamboreroOncodriveCLUST2013; @lawrenceDiscoverySaturationAnalysis2014;
@kumarStatisticallyIdentifyingTumor2015; @jiangWITER2019;
@zhaoModelingPositive2019; @noncoding-mutations-Zhang]. Hence, we sought to
simplify the problem by employing a "knowledge-base driven analysis"
[@pathways-Atul-Butte], investigating cancer as a disease of basic cellular and
biochemical pathways, which we accomplished by translating gene-level mutations
into pathway level disruptions. Our approach differs from previously described
methods [@PATHNET] in that we chose to focus on the pathway (defined in the
methods) as the unit of disruption instead of the gene, where individual
mutations may be sufficient to disrupt or misdirect pathway activity. Rather
than evaluating the burden of mutations on each pathway within or across tumors,
which can be biased by the number of genes or the coverage of pathway-linked
genes in the genome, we scored each pathway in binary fashion, allowing one
mutation or multiple mutations to count equally for each sample.

To our knowledge, this approach has not previously been attempted in spite of
its relative simplicity. We limited our analysis to mutations with likely
deleterious effects, including missense, nonsense, frame-shift, stop-loss,
untranslated region and splice site mutations in genes that are actively
expressed in each cancer type, thus avoiding bias from transcription coupled
repair. Our method of filtration differs from the "rank-and-cut" method of
[@PCAWG-MAIN] but represents a reasonable attempt to account for the same
biases. We also filtered genes on the basis of pathway membership in biochemical
pathways, excluding curated gene sets related to diseases, syndromes, or classes
of proteins with shared catalytic activity or conserved domains which are
potentially problematic [@pathways-Atul-Butte]. We chose this approach to limit
redundancy and exclude biologically unrelated collections of genes. Our
pathway-centric analysis allows us to circumvent the problem that some
cancer-promoting mutations likely occur at such low frequencies as to be
indistinguishable from background.

The hypothesis that cancer is the result of dysfunction from a limited number of
basic cellular processes common to all living human cells was introduced and
later expanded on in a pair of essays by Hanahan and Weinberg [@weinberg2000;
@hallmarks_II]. In this view, all tumors regardless of tissue of origin acquire
six essential phenotypes; sustaining proliferative signaling, evading growth
suppressors, replicative immortality, evading apoptosis, supportive
angiogenesis, tissue invasion and metastasis [@weinberg2000]. Advancements in
the field over the ensuing decade led to the inclusion of new tumor promoting
processes, namely deregulation of cellular energy cycle, tumor promoting
inflammation, evasion of the immune system, genome instability, and on top of
these acquired traits a new appreciation for the concept of the tumor
microenvironment and its contribution to the evolution and promotion of
metastatic disease [@hallmarks_II].

An alternative hypothesis is that every tumor belongs to one of a large number
of syndromes which are unique to each tissue-of-origin, that share some
mechanisms and treatment strategies. Recent publication of TCGA consortium
papers present a view largely, and surprisingly, consistent with this hypothesis
[@hoadleyMultiplatformAnalysis2014; @hoadleyCellofOriginPatternsDominate2018].
Perhaps owing to the intractable complexity of genomics, proteomics, and
patient metadata in all its forms, the inescapable conclusion thus far is that
tissue-of-origin remains the most important driver of tumor characteristics at
every scale and by every measure. Our observations contrast with this view, and
instead support an interpretation of publicly available data in which all tumors
manifest one of a limited number of metastable phenotypes resulting from
disruptions in basic pathways as represented by our clusters.

We attempted to account for our clusters in terms of more trivial explanations.
For example, it could be that the clusters we identified are consistent with a
signature of overall disease progression, such that successive, adjacent
clusters on the UMAP projection exhibit increasing specificity. We identified no
such trend in the number of mutations, the relative staging or metastasis; and
each cluster instead was associated with unique combinations of pathways.

We find that most tumors segregate into one of ten biologically distinct
clusters, regardless of the cancer type. Some cancer types are unevenly
distributed, though we could not identify any cancers that were exclusive to a
single tumor cluster. Only cancer types with the fewest samples were found to be
absent from one or more clusters at all. Somewhat surprisingly to us, this
finding extends to histological subtypes of breast, head and neck cancers,
leukemias, _etc._ This result implies that many, if not most, histological
subtypes could reflect differences in cell-of-origin, rather than fundamental
differences in proximal cause. The four major subtypes of breast cancer
correspond to histological and molecular expression profiles that define them
and how they respond to experimental stimuli [@breast-subtypes2]. It has been
hypothesized that differences in the molecular regulators of development in the
precursor cell types present in the breast epithelium drive the various
histological phenotypes [@breast-subtypes; @breast-subtypes3]. Consistent with
this view we found that breast tumor samples of the Luminal A subtype were
heavily biased toward membership in clusters 1 and 7, and basal tumor samples
were biased toward cluster 6. However, both subtypes also contained samples in
every other cluster (without exception), and Luminal B and Her2 positive samples
are distributed across clusters. We interpret these data to mean that inherited
cell-of-origin molecular profiles may predispose certain precursor cell types
within the breast epithelium to forming tumors of one cluster or another but are
not determinative; this view is compatible with the previously stated hypothesis
but opens the way for a more granular view of individual tumors. It would be
surprising if we did not observe bias for some cancers and subtypes amongst our
classes, since some treatment regimens have greater efficacy for patients of a
given cancer or histological subtype [@breast-subtypes2]. Nonetheless, the basis
for some tumors being treatment-refractory in spite of receiving the standard of
clinical care for diagnostic markers remains elusive. Doubtless some of this is
due to chance events, as tumors can metastasize and remain dormant years before
they are detected at distal sites, or resistant clones may have already arisen
at undetectable levels [@hallmarks_II], but our analysis suggests the
possibility of identifying more informative molecular, histological or cellular
subtypes that could form a basis for future stratification of patients into
different precision treatment regimens.

@@

Our results illustrate how unique combinations of mutations in pan-cancer driver
genes with tissue-specific pathway disruptions result in common categories when
viewed at the level of the pathway knowledge-base. Top cancer driver genes (_e.g.
PIK3CA_ and _TP53_) are found in most of the clusters, in spite of the fact that
they contribute to many cluster-specific pathways. This can only be explained as
a result of other less common driver genes complementing the pathway disruption
of the unique combination of driver genes that are disrupted in each tumor, and
we speculate that many of these are relatively tissue-specific in terms of their
sensitivity to mutation. Consistent with our observations, @moonlight showed how
tissue- and cancer-specific driver genes and druggable targets of pathway moduli
can be discovered by machine learning on integrated multi-omics datasets.
Marrying these approaches in future work will greatly enhance our understanding
of tumor biology.

@@

Finally, we must remark on the limitations of our work exemplified in two of the
clusters, 1 and 10, for which we did not find many distinctive associations with
pathways. Cluster 1 had a relatively low proportion of mutated pathways across
the board, although it is broadly enriched in many of the same tumor-promoting
pathways common to the other groups. This likely does not reflect a true
difference in stages of tumor evolution, as our data show clearly that this
cluster is as likely to contain stage IV metastatic tumors as it is to contain
those of stage I. This cluster likely represents a group of tumors with
aberrations in methylation, key deletions, or other structural variants.
Consistent with this, kidney chromophobe and thyroid cancers have high
proportions of structural variations _vs._ other variant types [@PCAWG-MAIN] and
are heavily skewed to cluster 1 membership. Likewise, cluster 10 represents a
group of hyper-mutated tumors that harbor so many mutations that virtually no
pathway is unaffected. It seems likely that a significant fraction of the
“mutant” samples for each pathway are burdened with excess passenger mutations.
This could be addressed with application of ever-increasingly sophisticated
filtering of likely passenger mutations [_e.g._ @spliceAI;
@deep-learning-clinical-variants]. In the future, we hope to incorporate these
other data into a comprehensive pathway-centric analysis as we have done here
for point mutations and indels.

@@

By modeling CDC longevity data as a baseline risk function we showed that each
of our pathway disruption clusters exhibit cancer-type specific effects on
expected survival. Contrary to our initial expectation, we found that models
assuming different survival based on cluster $\times$ cancer type fit better than
models in which they are independent factors. However, considering that within
each cancer type there are different clinical standards of care, and even within
classes of drugs the preferred treatment can very between cancers, it makes
sense that we observe tissue-specific cluster effects. Contrast the situation
with ovarian _vs._ breast cancer, which are both hormonally driven cancers, for
example. Ovarian cancer has but one main treatment axis, platinum, whereas
breast cancer patients have a variety of treatment regimens based on molecular
subtype and other factors. Standard of care for one cancer and cluster could
produce different outcomes than the same cluster of a different cancer type.
Unfortunately, given the diversity of drug classes and treatments (including
non-chemical treatments such as surgery and palliative care), we lack sufficient
power to explore these variables in the TCGA data. It is our hope that future
studies will help to distinguish between treatment-specific effects on survival
given different pathway disruption clusters.

@@

Our findings imply that there are separate processes in the etiology of cancer
that can be broadly thought of as general cancer promoting, cluster-specific
mutations and metastasis. General cancer promoting processes must include
factors that relate to genome stability and immortality, as "enabling
characteristics" of the cancer phenotype [@hallmarks_II]. Such pathways are
disrupted in all or most of the defined clusters
and are frequently the result of aberrations involving common driver genes such
as BRCA1/2, MMR genes, mitotic checkpoints, cohesion complexes, _etc._
Cluster-specific
evolution must involve the acquisition of disruptions to pathways that may
individually be harmful ( _e.g._ highly proliferative cells are more likely to
senesce) but together produce more specialized cancer phenotype and increased
fitness. Importantly, our
observations do not imply the order in which these mutations should accumulate.
This could be addressed in a future study by integrating our analysis with
evolutionary analysis of clonality, drawing inferences from variant allele
frequencies as in @PCAWG-EVO. However, since many of the genes in the
non-cluster-specific pathways involve the known driver genes, it is reasonable
to hypothesize that these mutations promote or enable the subsequent acquisition
of cluster-specific defects via random mutation and natural selection, thus
producing the clusters we observed. In support of this, the pan-cancer analysis
of whole genomes consortium (PCAWG) found that oncogenic driver mutations are
highly enriched in early arising clones, whereas later arising clones have much
greater diversity in driver mutations [@PCAWG-EVO]. Moreover, driver genes that
are known to be responsible for discrete mutation signatures such as APOBEC,
BRCA1 and BRCA2 produce mutational hotspots reflecting the varying selective
pressures in different tissues [@PCAWG-MAIN].

@@

We report that pathway enrichment in metastatic tumors across all clusters
(pooled) yielded generally lower effect sizes and larger p-values than the
cluster specific analysis, suggesting that signal is diluted when the samples
are analyzed together, and supporting the view that metastasis has
cluster-specific requirements. Since the number of metastatic samples is
relatively low, this part of our analysis is likely underpowered and subject to
expanded analysis with larger cohorts. The fact that most metastatic enrichment
is cluster-specific and has a tendency to overlap with cluster-specific pathways
from non-metastatic tumors of neighboring clusters suggests a mechanism of
complementarity, in which newly acquired mutations result in similar clusters
converging on one or more deadly phenotypes that capture most of the critical
features of end-stage cancer. Thus, even at relatively low power, our analysis
of metastasis further supports the pathway-centric approach as a fruitful way to
uncover biologically meaningful differences between tumors against the noisy
backdrop of tissue specific profiles. As a major caveat to this, we acknowledge
that many therapies _do_ target highly-specific driver genes, markers, and
signaling pathways ( _e.g._ TP53, EGFR or HER2), but understanding the broader
context of the genetic background and pathway vulnerability of tumors containing
such markers may aid in creating smarter combination therapies. We submit that
when we discover the requirements of each cluster with respect to pathway
disruptions and metastasis we may be able to target them therapeutically and
prevent further adaptation.

>>>>

One of the biggest hurdles in cancer research is the sparsity of data; ~20,000
protein-coding genes is comparable with the number of tumor samples, even with
multiple mutations per sample. We sought to
simplify the problem by employing a "knowledge-base driven analysis"
[@pathways-Atul-Butte], investigating cancer as a disease of basic cellular and
biochemical pathways, which we accomplished by translating gene-level mutations
into pathway level disruptions. Our approach differs from previously described
methods [@PATHNET] in that we chose to focus on the pathway (defined in the
methods) as the unit of disruption instead of the gene, where individual
mutations may be sufficient to alter pathway activity.

To our knowledge, this approach has not previously been attempted despite
its relative simplicity. We limited our analysis to mutations with likely
deleterious effects, including missense, nonsense, frame-shift, stop-loss,
untranslated region and splice site mutations in genes that are actively
expressed in each cancer type, thus avoiding bias from transcription coupled
repair. Our method of filtration differs from the "rank-and-cut" method of
[@PCAWG-MAIN] but represents a reasonable attempt to account for the same
biases. We also restricted our analysis to biochemical
pathways, excluding curated gene sets related to diseases, syndromes, or classes
of proteins with shared catalytic activity or conserved domains which are
potentially problematic [@pathways-Atul-Butte]. We chose this approach to limit
redundancy and exclude biologically unrelated collections of genes.

The hypothesis that cancer is the result of dysfunction from a limited number of
basic cellular processes common to eukaryotic cells was introduced and
later expanded on in a pair of essays by Hanahan and Weinberg [@weinberg2000;
@hallmarks_II]. An alternative hypothesis is that every tumor belongs to one of a large number
of syndromes which are unique to each tissue-of-origin, that share some
mechanisms and treatment strategies. Recent publication of TCGA consortium
papers present a view largely, and surprisingly, consistent with this latter hypothesis
[@hoadleyMultiplatformAnalysis2014; @hoadleyCellofOriginPatternsDominate2018].
Perhaps owing to the intractable complexity of genomics, proteomics, and
patient metadata in all its forms, the inescapable conclusion thus far is that
tissue-of-origin remains the most important driver of tumor characteristics at
every scale and by every measure. Our observations contrast with this view, and
instead support an interpretation of publicly available data in which all tumors
manifest one of a limited number of phenotypes resulting from
disruptions of basic pathways.

We attempted to account for our clusters in terms of more trivial explanations.
For example, it could be that the clusters we identified are consistent with
disease progression. We were unable to identify any such trend in the number of
mutations, the relative staging or metastasis; and each cluster instead was
associated with unique combinations of pathways.

Some cancer types are unevenly
distributed among the clusters, though we could not identify any cancers that were exclusive to a
single cluster. Only cancer types with the fewest samples were found to be
absent from one or more clusters at all. Somewhat surprisingly to us, this
finding extends to histological subtypes of breast, head and neck cancers,
leukemias, _etc._ This result implies that histological
subtypes could reflect differences in cell-of-origin, rather than fundamental
differences in cancer phenotype. The four major subtypes of breast cancer
correspond to histological and molecular expression profiles that define them
and how they respond to experimental stimuli [@breast-subtypes2]. It has been
hypothesized that differences in the molecular regulators of development in the
precursor cell types present in the breast epithelium drive the various
histological phenotypes [@breast-subtypes; @breast-subtypes3]. Consistent with
this view we found that breast tumor samples of the Luminal A subtype were
heavily biased toward membership in clusters 1 and 7, and basal tumor samples
were biased toward cluster 6. However, both subtypes also contained samples in
every other cluster (without exception), and Luminal B and Her2 positive samples
are distributed across clusters. We interpret these data to mean that inherited
cell-of-origin signatures could predispose certain precursor cells
within the breast epithelium to forming tumors of one cluster or another but are
not determinative. This view is compatible with the previously stated hypothesis
but opens the way for a more granular view of individual tumors. It would be
surprising if we did not observe bias for some cancers and subtypes amongst our
classes, since some treatment regimens have greater efficacy for patients of a
given cancer or histological subtype [@breast-subtypes2]. Nonetheless, the basis
for some tumors being treatment-refractory in spite of receiving the standard of
clinical care for diagnostic markers remains elusive. Doubtless some of this is
due to chance events, as tumors can metastasize and remain dormant years before
they are detected at distal sites, or resistant clones may have already arisen
at undetectable levels [@hallmarks_II], but our analysis suggests the
possibility of identifying more informative molecular, histological or cellular
subtypes that could form a basis for future stratification of patients into
different precision treatment regimens.

@@

Our results illustrate how unique combinations of mutations in pan-cancer driver
genes with tissue-specific pathway disruptions result in common categories when
viewed at the level of the pathway knowledge-base. Top cancer driver genes (_e.g.
PIK3CA_ and _TP53_) are found in most of the clusters, in spite of the fact that
they contribute to many cluster-specific pathways. This can only be explained as
a result of other less common driver genes complementing the pathway disruption
of the unique combination of driver genes that are disrupted in each tumor, and
we speculate that many of these are sensitive to mutation in different tissues.

@@

Finally, we must remark on the limitations of our work exemplified in two of the
clusters, 1 and 10, for which we did not find many distinctive associations with
pathways. Cluster 1 had a relatively low proportion of mutated pathways,
although it is broadly enriched in many of the same tumor-promoting pathways
common to the other groups. In addition, our data show clearly that this cluster
is as likely to contain stage IV metastatic tumors as it is to contain those of
stage I. This cluster likely represents a group of tumors with aberrations in
methylation, copy number, or other structural variants. Consistent with this,
kidney chromophobe and thyroid cancers have high proportions of structural
variations _vs._ other variant types [@PCAWG-MAIN] and are heavily skewed to
cluster 1 membership. Likewise, cluster 10 represents a group of hyper-mutated
tumors that harbor so many mutations that virtually no pathway is unaffected. It
seems likely that a significant fraction of the “mutant” samples for each
pathway are burdened with excess passenger mutations. This could be addressed
with more sophisticated filtering of likely passenger mutations [_e.g._
@spliceAI;
@deep-learning-clinical-variants]. In the future, we hope to incorporate these
other data into a comprehensive pathway-centric analysis as we have done here
for point mutations and indels.

@@

By modeling CDC longevity data as a baseline risk function we showed that each
of our pathway disruption clusters exhibit cancer-type specific effects on
expected survival. However, considering that within each cancer type there are
different clinical standards of care, and even within classes of drugs the
preferred treatment can very between cancers, it makes sense that we observe
tissue-specific cluster effects. Contrast the situation with ovarian _vs._
breast cancer, which are both hormonally driven cancers, for example. Ovarian
cancer has but one main treatment axis, platinum, whereas breast cancer patients
have a variety of treatment regimens based on molecular subtype and other
factors. Unfortunately, given the diversity of drug classes and treatments, we
lack sufficient power to explore these variables in the TCGA data. It is our
hope that future studies will help to distinguish between treatment-specific
effects on survival given different pathway disruption clusters.

@@

Our findings imply that there are separate processes in the etiology of cancer
that can be broadly thought of as general cancer promoting, cluster-specific
mutations and metastasis. General cancer promoting processes include genome
stability and immortality, as "enabling characteristics" of the cancer phenotype
[@hallmarks_II]. Such pathways are disrupted in most clusters and are frequently
the result of aberrations involving common driver genes such as BRCA1/2, MMR
genes, mitotic checkpoints, cohesion complexes, _etc._ Cluster-specific
evolution must involve the acquisition of disruptions to pathways that may
individually be harmful (_e.g._ highly proliferative cells are more likely to
senesce) but together produce more specialized cancer phenotype and increased
fitness. Importantly, our observations do not imply the order in which these
mutations should accumulate. This could be addressed in a future study by
evolutionary analysis of clonality, drawing inference from variant allele
frequencies as in @PCAWG-EVO. However, since many of the genes in the
non-cluster-specific pathways involve the known driver genes, it is reasonable
to surmise that these mutations promote or enable acquisition of
cluster-specific defects via random mutation and natural selection, thus
producing the clusters we observed. In support of this, the pan-cancer analysis
of whole genomes consortium (PCAWG) found that oncogenic driver mutations are
highly enriched in early arising clones, whereas later arising clones have much
greater diversity in driver mutations [@PCAWG-EVO]. Moreover, driver genes that
are known to be responsible for discrete mutation signatures such as APOBEC,
BRCA1 and BRCA2 produce mutational hotspots reflecting varying selective
pressures in different tissues [@PCAWG-MAIN].

@@

We report that enrichment in metastatic tumors across all clusters yielded
generally lower effect sizes and larger p-values than the cluster specific
analysis, suggesting that signal is diluted when clusters are pooled, and
supporting the view that metastasis has cluster-specific requirements. Since the
number of metastatic samples is relatively low, this part of our analysis is
likely underpowered and subject to expanded analysis with larger cohorts. The
fact that most metastatic enrichment is cluster-specific and has a tendency to
overlap with cluster-specific pathways from non-metastatic tumors of neighboring
clusters suggests that newly acquired mutations result in similar clusters
converging on one or more deadly phenotypes with critical features of end-stage
cancer. Thus, even at relatively low power, our analysis of metastasis uncovered
differences between tumors against the noisy backdrop of tissue specific
profiles. As a major caveat to our pan-cancer analysis, we acknowledge that many
therapies _do_ target highly-specific driver genes, markers, and signaling
pathways ( _e.g._ TP53, EGFR or HER2), but understanding the broader context of
the genetic background and pathway vulnerability of tumors containing such
markers may aid in creating smarter combination therapies. We submit that when
we discover the requirements of each cluster with respect to pathway disruptions
and metastasis we may be able to target them therapeutically and prevent further
adaptation.

### 2020-08-28 text/introduction.Rmd
cut all text to conform to Cell's 45k character limit

Recent advances in technology have greatly expanded our view into the mechanisms
of cancer at the systems biology level. Array and next-generation sequencing
technologies have made it possible to assay germline and somatic mutations,
expression profiles, DNA methylation, and copy number variations from the same
tissue samples. To take advantage of these tools, several large consortia,
including the cancer genome atlas (TCGA) and pan-cancer analysis of whole
genomes (PCAWG) sequenced large numbers of tumors and collected data from
multiple assays that could be analyzed together, with the goal of integration
and increased understanding of the mechanisms of cancer.

Considerable progress has been made analyzing these data. Statistical analyses
identify hundreds of global and tissue-specific cancer driver genes
[@deesMuSiC2012; @tamboreroOncodriveCLUST2013;
@lawrenceDiscoverySaturationAnalysis2014;
@kumarStatisticallyIdentifyingTumor2015; @tokheimEvaluatingEvaluationCancer2016;
@jiangWITER2019; @zhaoModelingPositive2019] using approaches aimed at detecting
when genes are mutated at a greater rate than expected due to chance. It has
been estimated that fewer than five mutations in key oncogenes and/or tumor
suppressors would be sufficient to transform a normal cell to a cancerous state
[@vogelsteinPathCancerThree2015; @Iranzo-Network]. Somewhat paradoxically,
mutations in cancer driver genes are commonplace in healthy tissue and
correlate with age and environmental exposures [@mutations-healthy-tissues].

Other studies provide a comprehensive view of mutations, gene expression and
genomic signatures, with the express goal of understanding the common themes of
all cancer independent of tissue of origin. Understanding cancer as a disease of
the cell has long been a goal of the field as characterized in essays by
@weinberg2000; @hallmarks_II. The first of these studies in the genomic era
identified 11 subtypes from 12 cancer types, using integrative analysis with
co-equal weighting of gene expression, methylation, copy number and proteomics
data [@hoadleyMultiplatformAnalysis2014]. The principal finding was that
tissue-of-origin is the predominant driving factor, though a significant
proportion of tumors (~10%) could be reclassified independent of
tissue-of-origin. In a second study involving 33 cancer types and a much greater
number of tumors, the authors identify
28 clusters that could be further subdivided into organ specific groups,
including pan-gastrointestinal, pan-gynecological, pan-squamous,
pan-gynecological/squamous and pan-kidney [@hoadleyCellofOriginPatternsDominate2018].

More recently, attempts have been made to consider pan-cancer mutations and
other data as disruptions of normal pathway activity. @PCAWG-PATH carried out
near-exhaustive analysis of pathway enrichment using whole genome data. Network
approaches address the sparseness of mutations, allowing genes to be influenced
by mutations in the nearest network neighbors
[@hornNetSigNetworkbasedDiscovery2018; @Iranzo-Network]. These methods are
powerful especially when applied to a small number (on the order of hundreds) of
tumors. One of the challenges in the field remains that beyond a couple hundred
common driver genes and mutations, there may be thousands of moderate effect
genes that occur at such low frequency--given tissue diversity--and low sample
numbers that they are impossible to detect using positive selection theory. Some
groups have attempted to address this challenge by applying machine learning to
cancer data sets to discover groups of functionally related genes as they
interact with larger pathways and networks [@kimPancancerAnalysisSomatic2018;
@mourikis-ml-pathways; @moonlight]. These studies have proven very effective at
highlighting fundamental disease phenotypes at the pathway
level across cancers with different origins at the cellular and tissue
level.

>>>>

Recent advances in technology have greatly expanded our view into the mechanisms
of cancer at the systems biology level. Next-generation sequencing technologies
have made it possible to assay germline and somatic mutations, expression
profiles, DNA methylation, and copy number variations from the same tissue. To
take advantage of these tools, several large consortia, including the cancer
genome atlas (TCGA) and pan-cancer analysis of whole genomes (PCAWG) sequenced
large numbers of tumors multiple assays to analyze together, with the goal of
integration and increased understanding of the mechanisms of cancer.

Considerable progress has been made analyzing these data. Statistical analyses
identify hundreds of global and tissue-specific cancer driver genes
[@deesMuSiC2012; @tamboreroOncodriveCLUST2013;
@lawrenceDiscoverySaturationAnalysis2014;
@kumarStatisticallyIdentifyingTumor2015; @tokheimEvaluatingEvaluationCancer2016;
@jiangWITER2019; @zhaoModelingPositive2019] using approaches aimed at detecting
genes mutated at a greater rate than expected due to chance. It has been
estimated that fewer than five mutations in key oncogenes and/or tumor
suppressors would be sufficient to transform a normal cell to a cancerous state
[@vogelsteinPathCancerThree2015; @Iranzo-Network]. Mutations in cancer driver
genes are commonplace in healthy tissue and correlate with age and environmental
exposures [@mutations-healthy-tissues].

Other studies provide a comprehensive view of mutations, gene expression and
genomic signatures, with the goal of understanding common themes of all cancer
independent of tissue of origin. Understanding cancer as a disease of the cell
has long been a goal of the field as characterized in essays by
@weinberg2000; @hallmarks_II. The first of these studies in genomics identified
11 subtypes from 12 cancer types, using integrative analysis with co-equal
weighting of gene expression, methylation, copy number and proteomics data
[@hoadleyMultiplatformAnalysis2014]. The principal finding was that
tissue-of-origin is the predominant driving factor, though ~10% could be
reclassified independent of tissue-of-origin. In a second study involving 33
cancer types and a much greater number of tumors, the authors identified 28
clusters that could be further subdivided into organ specific groups, including
pan-gastrointestinal, pan-gynecological, pan-squamous,
pan-gynecological/squamous and pan-kidney
[@hoadleyCellofOriginPatternsDominate2018].

More recently, pan-cancer mutations and other data have been looked at as
disruptions of normal pathway activity. @PCAWG-PATH carried out exhaustive
analysis of pathway enrichment using whole genome data. Network approaches
address the sparseness of mutations, allowing genes to be influenced by
mutations in nearest network neighbors [@hornNetSigNetworkbasedDiscovery2018;
@Iranzo-Network]. These methods are powerful especially when applied to smaller
numbers of tumors. One of the challenges in the field remains that there may be
thousands of moderate effect genes that occur at such low frequency--given
tissue diversity and low sample number--that they are impossible to detect using
positive selection theory. Some groups have attempted to address this challenge
by applying machine learning to cancer data to discover groups of
functionally related genes as they interact with larger pathways and networks
[@kimPancancerAnalysisSomatic2018;
@mourikis-ml-pathways; @moonlight]. These studies have proven very effective at
highlighting fundamental disease phenotypes at the pathway level across cancers
with different origins at the cellular and tissue level.
